---
title: "Project Title"
subtitle: "Project subtitle (if applicable)"
author: "Your names | Group #X"
date: today
format:
  beamer:
    theme: Madrid
    colortheme: default
    fonttheme: serif
    aspectratio: 169
    slide-level: 2
    toc: false
    incremental: false
    pdf-engine: pdflatex
    keep-tex: false
editor: visual
---

```{r setup, include=FALSE}
library(gridExtra)
library(tidyverse)
library(knitr)
library(cowplot)
knitr::opts_chunk$set(
  echo = FALSE,
  include = TRUE,
  warning = FALSE,
  message = FALSE,
  results = "markup",
  fig.width = 6,
  fig.height = 3.5,
  fig.align = "center",
  out.width = "80%"
)
```

```{r load-data, echo=FALSE}
### Read in data
CATF_data <- read_csv("CATF_data.csv")
```

```{r clean_data, include=FALSE}

## Clean the data

## E.g., remove unrealistic populations

#CATF_data |> arrange(population) |> View()

#CATF_data <- CATF_data |> filter(population > 100) ## Alternatively, we could convert them to the correct values for each country by ## seeing roughly what the population values for each country are and scaling up accordingly.

# ============================================================

# Aggregate CATF_data: one row per country

# ============================================================

# Each country has multiple rows (one per student). We collapse

# them into a single row using sensible summary functions.

# 

# General strategy:

# - numeric columns → median (robust to outliers / typos)

# - character columns → mode (most common value reported)

# 

# Feel free to swap median() for mean() or mode() on any

# variable if you think a different choice is more appropriate.

# ============================================================

# --- helper function: statistical mode (most frequent value) ----

# works for both character and numeric vectors

stat_mode <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA)
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# --- identify column types --------------------------------------

numeric_cols <- CATF_data |> select(where(is.numeric)) |> names()

character_cols <- CATF_data |> select(where(is.character), -country) |> names()

# --- aggregate -----------------------------------------------

CATF_clean <- CATF_data |> group_by(country) |> summarise( # For every numeric column, take the median (ignoring NAs) across(all_of(numeric_cols), function(x){ median(x, na.rm = TRUE) }),

# For every character column, take the mode
across(all_of(character_cols), stat_mode),

# (optional) track how many student rows were combined
n_sources = n()
)

# --- quick check ---------------------------------------------

#glimpse(CATF_clean)

# How many unique countries do we have?

#nrow(CATF_clean)

# ============================================================

# CUSTOMISATION IDEAS

# ============================================================

# If you'd prefer to use the mean for some variables and the

# median for others, you can split them up like this:

# 

# CATF_clean <- CATF_data |>

# group_by(country) |>

# summarise(

# # Variables where mean makes sense

# across(c(gdp, population),

# function(x){ mean(x, na.rm = TRUE) } ),

# 

# # Variables where median is more robust

# across(c(poverty_rate, hdi, median_age),

# function(x){ median(x, na.rm = TRUE) } ),

# 

# # Variables where mode makes sense (e.g. categorical-ish)

# across(c(renewable_target_solar, renewable_target_wind),

# stat_mode),

# 

# n_sources = n()

# )

# ============================================================

```

Introduction

Overall Theme: [Patterns and drivers of renewable energy development across African electricity systems]

Research Goal: [Investigate how renewable energy usage is different among African electricity systems and explore factors that may be associated with differences in renewable energy capacity between countries.]

Target Population: [The study focuses on African countries included in the CATF electricity sector dataset.]

Audience: [The primary audience is Clean Air Task Force (CATF) staff, who are interested in understanding patterns in electricity sector performance across African countries to help inform the development of the Electricity Sector Performance Scorecard (CATF-ESPS).]

Data Summary

The dataset contains country-level indicators describing electricity system performance across African countries. These indicators capture areas such as electricity access, reliability, infrastructure development, financial performance, and renewable energy capacity.

Key Variables:

Variable 1: [Description]

Variable 2: [Description]

Data Wrangling:

[- To facilitate comparison across countries, the variable `collection_rate_percent` was grouped into categories representing different levels of electricity revenue collection performance. A new variable called `collection_group` was created in the `CATF_data` dataset. The groups are defined as Low (<70%), Medium (70–90%), and High (>90%). This transformation allows us to compare `total_installed_variable_renewables_capacity` across countries with different levels of revenue collection performance.
]

## Question #1: [Caellum Yip Hoi - Lee - Renewable Presence and Power Reliability]
Do countries with a higher share of renewable energy in the generation mix tend to experience more frequent power interruptions?

## Q1: Variables & Approach
**Variables used:**

- **renewables_share_gen_mix_percent:**
  The percentage of electricity generation coming from renewable energy sources.
  This variable captures the amount of renewable energy being employed in the country's electricity system.

- **SAIFI:**
  System Average Interruption Frequency Index, representing Voltage Quality / Frequency 
  It is the total number of reported interruptions divided by total customers. This variable serves as a primary measure of power reliability.
  
**New variables/groups defined:**

To intuitively compare reliability, I can group countries based on their renewable generation share:

- renewable_group: "Low" when renewables_share_gen_mix_percent is below 20%, "Medium" when it is between 20-50%, and "High" when above 50%
  
  
## Q1: Visualization
```{r q1-plot}
CATF_data |>
  # An extreme outlier of 65,000 % exists in the data, filtered this out because it makes no sense and is likely a typo. 
  filter(renewables_share_gen_mix_percent <= 100) |> 
  ggplot(aes(x = renewables_share_gen_mix_percent, y = SAIFI)) +
  geom_point(color = "steelblue", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    x = "Renewables Share in Generation Mix (%)",
    y = "Frequency of Interruptions (SAIFI)",
    title = "Renewable Energy Share vs Power Reliability"
  )
```

## Q1: Summary Table

```{r q1-table}
CATF_data |>
  # An extreme outlier of 65,000 % exists in the data, filtered this out because it makes no sense and is likely a typo. 
  filter(renewables_share_gen_mix_percent <= 100) |>
  mutate(
    renewable_group = case_when(
      renewables_share_gen_mix_percent < 20 ~ "Low (<20%)",
      renewables_share_gen_mix_percent <= 50 & renewables_share_gen_mix_percent >= 20 ~ "Medium (20–50%)",
      renewables_share_gen_mix_percent > 50  ~ "High (>50%)"
    )
  ) |>
  
  mutate(renewable_group = factor(renewable_group, levels = c("Low (<20%)", "Medium (20–50%)", "High (>50%)"))) |>
  group_by(renewable_group) |>
  summarise(
    Countries = n(),
    `Avg Renewable Share (%)` = round(mean(renewables_share_gen_mix_percent, na.rm = TRUE), 1),
    `Avg SAIFI (Interruptions)` = round(mean(SAIFI, na.rm = TRUE), 1)
  ) |>
  kable(caption = "Average System Interruptions (SAIFI) by Renewable Energy Share")
```

## Q1: Insights & Takeaways

* **No clear linear relationship:** The scatter plot has a nearly flat trend line, suggesting that a higher share of renewables in the generation mix is not correlated with a higher frequency of power interruptions.
* **Reliability driven by other factors:** Countries with high interruption frequencies had widely varying levels of renewable energy in their generation mix. This suggests that issues with reliability and the subsequent SAIFI score likely has more to do with other factors, i.e infrastructure. 
* **Renewables not sacrificing stability:** I see this evidence as positive, since it suggests that African electricity systems can have higher percentages of renewable energy without directly inducing a lower electric grid reliability.


Question #2: [Name - Question Statement]

[Repeat structure from Q1]

Q2: Variables & Approach

Q2: Visualization

Q2: Summary Table

## Question #3: [Name - Question Statement]

[Clear statement of the research question]

## Q3: Variables & Approach

Variables used:

[List variables]

New variables/groups defined:

[Any created variables]

## Q3: Visualization

```{r q3-plot}
#Your plot code here
```

## Q3: Summary Table

```{r q3-table}
#Your summary table here
```

## Insights for Q3


## Question #4: [Sahand Samadirad - Revenue Collection and Renewable Capacity]

Do countries with higher electricity revenue collection rates tend to have more installed renewable energy capacity?

## Q4: Variables & Approach

**Variables used:**

- **collection_rate_percent**\
  Percentage of billed electricity revenue that utilities successfully collect.\
  This variable reflects the financial efficiency and operational performance of electricity utilities.

- **total_installed_variable_renewables_capacity**\
  Total installed capacity of variable renewable energy sources such as wind and solar (MW).\
  This variable represents the level of renewable energy development within a country's electricity system.

**New variables/groups defined:**

To make comparisons clearer, countries are grouped based on electricity revenue collection performance:

- Low collection rate: below 70%\
- Medium collection rate: 70–90%\
- High collection rate: above 90%

These categories allow us to compare renewable capacity across different levels of financial performance.

## Q4: Visualization

```{r q4-plot}
CATF_data |>
  ggplot(aes(x = collection_rate_percent, y = total_installed_variable_renewables_capacity)) +
  geom_point(color = "darkgreen", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    x = "Electricity Revenue Collection Rate (%)",
    y = "Installed Renewable Energy Capacity (MW)",
    title = "Collection Rate vs Renewable Energy Capacity"
  )
```

## Q4: Summary Table

```{r q4-table}
CATF_data |>
  mutate(
    collection_group = case_when(
      collection_rate_percent < 70 ~ "Low (<70%)",
      collection_rate_percent < 90 ~ "Medium (70–90%)",
      TRUE ~ "High (>90%)"
    )
  ) |>
  group_by(collection_group) |>
  summarise(
    Countries = n(),
    `Avg Collection Rate (%)` = round(mean(collection_rate_percent, na.rm = TRUE), 1),
    `Avg Renewable Capacity (MW)` = round(mean(total_installed_variable_renewables_capacity, na.rm = TRUE), 1)
  ) |>
  kable(caption = "Average Renewable Capacity by Collection Rate Category")
```

## Overall Limitations

It is optional to include this as a stand-alone section at the end, you could also choose to discuss this at the end of each research question

Data Limitations:

[Limitation 1]

[Limitation 2]

Future Directions:

[How these could be addressed]

Overall Conclusion

Key Findings:

[Finding 1 related to theme]

[Finding 2 related to theme]

[Finding 3 related to theme]

Connection to Theme:

[How findings relate back to unifying theme]

References & Acknowledgements (if applicable)

References:

[Citation 1, if applicable]

Acknowledgements:

[Any acknowledgements, if applicable]
