---
title: "Renewable Energy Development and Electricity System Performance in Africa"
subtitle: "An Exploratory Analysis Using the CATF Electricity Sector Dataset"
author: "Caellum Yip Hoi Lee | Paanini Kartik | Hyman Al-Qaissi | Sahand Samadirad | \n Group C TUT0203"
date: today
format:
  beamer:
    theme: Madrid
    colortheme: default
    fonttheme: serif
    aspectratio: 169
    slide-level: 2
    toc: false
    incremental: false
    pdf-engine: pdflatex
    keep-tex: false
editor: visual
---

```{r setup, include=FALSE}
library(gridExtra)
library(tidyverse)
library(knitr)
library(cowplot)
library(broom)
knitr::opts_chunk$set(
  echo = FALSE,
  include = TRUE,
  warning = FALSE,
  message = FALSE,
  results = "markup",
  fig.width = 6,
  fig.height = 3.5,
  fig.align = "center",
  out.width = "80%"
)
```

```{r load-data, echo=FALSE}
### Read in data
CATF_data <- read_csv("CATF_data.csv")
```

## **Introduction** {shrink=10}

**Overall Theme:** Patterns and drivers of renewable energy development across African electricity systems

**Research Goal:** Investigate how renewable energy usage is different among African electricity systems and explore factors that may be associated with differences in renewable energy capacity between countries.

**Target Population:** The study focuses on African countries included in the CATF electricity sector dataset.

**Audience:** The primary audience is Clean Air Task Force (CATF), a global non-governmental organization that seeks to mitigate climate change through development and deployment of low-carbon energy technologies. CATF staff are interested in understanding patterns in electricity sector performance across African countries to help inform the development of the Electricity Sector Performance Scorecard (CATF-ESPS).

**Data Summary**

The dataset contains country-level indicators describing electricity system performance across African countries. These indicators capture areas such as electricity access, reliability, infrastructure development, financial performance, and renewable energy capacity.

## Question #1: \[Caellum Yip Hoi - Lee - Renewable Presence and Power Reliability\]

Do countries with a higher share of renewable energy in the generation mix tend to experience more frequent power interruptions?

* **Relevance:** As African nations develop electric grids, transitioning to renewable energy is a priority. Understanding how the shift to renewable sources could impact energy reliability may be helpful in leading effective infrastructure development. 
* **Connection to Theme:** This investigates a potential trade-off between two core areas of the CATF-ESPS: renewable energy development and power grid stability.

## Q1: Variables & Approach  {shrink=10}

**Variables used:**

-   **Renewable Energy Share Percent:** The percentage of electricity generation coming from renewable energy sources. This variable captures the amount of renewable energy being employed in the country's electricity system.

-   **SAIFI:** System Average Interruption Frequency Index, representing Voltage Quality / Frequency. It is the total number of reported interruptions divided by total customers. This variable serves as a primary measure of power reliability.

**New variables/groups defined:**

To intuitively compare reliability, we group countries based on their renewable generation share:

-   Renewable Group: "Low" when the countries renewable energy share is below 20%, "Medium" when it is between 20-50%, and "High" when above 50%

These cutoffs are established by the International Energy Agency (IEA). 20% marks the Phase 3 engineering tipping point for grid modernization, while 50% is the standard benchmark for renewables integration studies (IEA Wind Task 25 & PVPS Task 14).

## Q1: Visualization  {shrink=10}

```{r q1-plot}
country_level_data <- CATF_data |>
  group_by(country) |>
  summarise(
    country_renewable_share = median(renewables_share_gen_mix_percent, na.rm = TRUE),
    country_SAIFI = median(SAIFI, na.rm = TRUE)
  ) |>
  filter(!is.na(country_renewable_share) & !is.na(country_SAIFI)) |>
  # Anything with a percentage higher than 100 is impossible and likely a typo, so it can be filtered out. 
  filter(country_renewable_share <= 100)

q1_data <- country_level_data |>
  mutate(
    renewable_group = case_when(
      country_renewable_share < 20 ~ "Low (<20%)",
      country_renewable_share >= 20 & country_renewable_share <= 50 ~ "Medium (20–50%)",
      country_renewable_share > 50  ~ "High (>50%)"
    ),
    renewable_group = factor(renewable_group, levels = c("Low (<20%)", "Medium (20–50%)", "High (>50%)"))
  )

plot_zoomed <- ggplot(q1_data, aes(x = renewable_group, y = country_SAIFI, fill = renewable_group)) +
  geom_boxplot(alpha = 0.6, outlier.color = "red") +
  coord_cartesian(ylim = c(0, 150)) +
  scale_x_discrete(labels = c("Low\n(<20%)", "Medium\n(20–50%)", "High\n(>50%)")) +
  labs(
    x = "Renewable Share",
    y = "Interruptions (SAIFI)",
    title = "Core SAIFI Scores",
  ) +
  theme_minimal() +
  theme(legend.position = "none")

plot_full <- ggplot(q1_data, aes(x = renewable_group, y = country_SAIFI, fill = renewable_group)) +
  scale_x_discrete(labels = c("Low\n(<20%)", "Medium\n(20–50%)", "High\n(>50%)")) + 
  geom_boxplot(alpha = 0.6, outlier.color = "red") +
  labs(
    x = "Renewable Share",
    y = "", 
    title = "Complete SAIFI Range",
  ) +
  theme_minimal() +
  theme(legend.position = "none")

plot_grid(plot_zoomed, plot_full, ncol = 2)


```

## Q1: Summary Table  {shrink=10}

```{r q1-table}
q1_data |>
  group_by(renewable_group) |>
  summarise(
    Countries = n_distinct(country),
    `Avg Renewable Share (%)` = round(mean(country_renewable_share, na.rm = TRUE), 1),
    `Avg SAIFI` = round(mean(country_SAIFI, na.rm = TRUE), 1),
    `Max SAIFI` = round(max(country_SAIFI, na.rm = TRUE), 1),
    `Extreme Outages (>200 SAIFI)` = sum(country_SAIFI > 200, na.rm = TRUE)
  ) |>
  rename(`Renewable Group` = renewable_group) |>
  kable(caption = "System Interruptions and Extreme Outages by Renewable Energy Share")

```

## Q1: Insights & Takeaways  {shrink=10}

-   **Medium Power Share Stability:** Countries with a moderate renewable energy share (those with around 20-50%) exhibited the most stable and reliable grids. This group has the lowest average interruptions (13.6 SAIFI), the tightest visual spread, and zero extreme outliers, showing a strong association between a mixed generation profile and grid stability.
-   **Extreme Outliers in Low Group:** The Low (<20%) energy share group displayed the most widespread variance. While the median for the group is relatively low, it contains massive, catastrophic outliers (with SAIFI scores reaching over 600).
-   **High vs Low Integration: ** Countries in the High(>50%) renewable group had a slightly elevated median compared to the Medium and Low Groups, suggesting an overall increase in day-to-day fluctuations. However, the spread is significantly less than the low group, and the outliers are far less extreme. 
-   **Overall CATF Takeaway:** While this analysis cannot establish direct causation, it suggests that integrating renewable energy does not sacrifice stability. A mixed grid (20-50% renewables) performs the best, and transitioning to high renewables has more evidence of being safer than remaining in a low-renewable, traditional grid.

## Question #2: \[Paanini Kartik - Development and Renewable Energy Usage\]

**Do more developed countries tend to generate a higher or lower proportion of renewable energy?**

-  From one perspective, we may expect to see developed countries have a **higher proportion** of their energy generation capabilities come from renewable energy, given there is more capital available to invest and higher levels of education could mean more skilled workers capable to building renewable energy sources, while from another, we may expect to see developed countries have a **lower share in renewables**, as their higher development may correlate with the choice to use cheaper sources of energy such as oil and coal.

-  Understanding how development correlates with renewables is crucial to understanding electricty systems in Africa and how renewable usage is distributed in the continent.

## Q2: Variables & Approach

**Variables from data:**

-   **Human Development Index (HDI)** The Human Development Index of a given country. This will be used as a metric for how developed the country is, where HDI measures factors such as life expectancy, education, and GNI to assess how developed a country is.
-   **% of energy generated from renewables** What % of the country's generation comes from renewables. We use this instead of the total installed renewables capacity as that may trivially become higher with more development.

**Created variables:**

We will create a new variable by grouping countries based on their HDI using the standard detailed by the UNDP, which defines: "Low HDI" as less than 0.5, "Medium HDI" as between 0.5 and 0.699, "High HDI" as between 0.7 and 0.799, and "Very High HDI" as 0.8 and above.

## Q2: Visualization

```{r q2-plot}
CATF_data |>
  group_by(country) |>
  summarise(
    hdi = median(hdi, na.rm = TRUE),
    renewables_share_gen_mix_percent = mean(renewables_share_gen_mix_percent, na.rm = TRUE)
  ) |>
  filter(!is.na(hdi), !is.na(renewables_share_gen_mix_percent)) |>
  filter(renewables_share_gen_mix_percent <= 100 &
         renewables_share_gen_mix_percent >= 0) |>
  filter(hdi >= 0 & hdi <= 1) |> # remove invalid values
  ggplot(aes(x = hdi, y = renewables_share_gen_mix_percent)) +
  geom_point(color = "darkred", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    x = "Human Development Index (HDI)",
    y = "Renewables Share of Generation Mix (%)",
    title = "Development Level vs Renewable Energy Usage"
  )
```

## Q2: Summary Table and Linear Model {shrink=10}

```{r q2-table}
CATF_data |>
  group_by(country) |>
  summarise(
    hdi = median(hdi, na.rm = TRUE),
    renewables_share_gen_mix_percent = mean(renewables_share_gen_mix_percent, na.rm = TRUE)
  ) |>
  filter(!is.na(hdi), !is.na(renewables_share_gen_mix_percent)) |>
  filter(renewables_share_gen_mix_percent <= 100 &
         renewables_share_gen_mix_percent >= 0) |>
  filter(hdi >= 0 & hdi <= 1) |> # remove invalid values
  mutate(
    hdi_grouping = case_when(
      hdi < 0.5 ~ "Low (<0.5)",
      hdi >= 0.5 & hdi < 0.7 ~ "Med. (0.5–0.7)",
      hdi >= 0.7 & hdi < 0.8 ~ "High (0.7–0.8)",
      hdi >= 0.8 ~ "Very High (>=0.8)"
    )
  ) |>
  mutate(
    hdi_grouping = factor(
      hdi_grouping,
      levels = c("Low (<0.5)", "Med. (0.5–0.7)", "High (0.7–0.8)", "Very High (>=0.8)")
    )
  ) |>
  group_by(hdi_grouping) |>
  summarise(
  Countries = n(),
    `Avg HDI` = round(mean(hdi, na.rm = TRUE), 2),
    `HDI SD` = round(sd(hdi, na.rm = TRUE), 2),
    `Avg Renewable Share (%)` = round(mean(renewables_share_gen_mix_percent, na.rm = TRUE), 1),
    `Renewable Share SD` = round(sd(renewables_share_gen_mix_percent, na.rm = TRUE), 1)
) |> rename(`HDI Group` = hdi_grouping) |>
     kable(caption = "Renewable Energy Share by HDI Development Group")
```

```{r q2-lm}
lm_q2 <- CATF_data |>
  group_by(country) |>
  summarise(
    hdi = median(hdi, na.rm = TRUE),
    renewables_share_gen_mix_percent = mean(renewables_share_gen_mix_percent, na.rm = TRUE)
  ) |>
  filter(!is.na(hdi), !is.na(renewables_share_gen_mix_percent)) |>
  filter(renewables_share_gen_mix_percent <= 100 &
         renewables_share_gen_mix_percent >= 0) |>
  filter(hdi >= 0 & hdi <= 1) |>
  lm(renewables_share_gen_mix_percent ~ hdi, data = _) |>
  tidy()

tibble(
  Term = c("Intercept", "Slope (HDI)"),
  Estimate = c(
    lm_q2 |> filter(term == "(Intercept)") |> select(estimate) |> pull(),
    lm_q2 |> filter(term == "hdi") |> select(estimate) |> pull()
  ),
  `p-Value` = c(
    lm_q2 |> filter(term == "(Intercept)") |> select(p.value) |> pull(),
    lm_q2 |> filter(term == "hdi") |> select(p.value) |> pull()
  )
) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable(caption = "Linear Model: Renewables Share ~ HDI", align = "c")
```

## Q2: Insights & Takeaways

-   Visually speaking, there appears to be a **slight negative correlation** between the two variables, given the line of best fit having a slightly negative slope.
-   However, there appears to be a large spread in the data, with data points not clustering around said line. The spread is more pronounced for countries with lower HDI scores, with their standard deviations being between 25-30, while countries with higher HDI scores have standard deviations less than 20.
-   Morover, the **p-value of 0.142** we get for the slope is above the common alpha threshold of 0.05, which means **we have do not have enough evidence to reject the null hypothesis** and we cannot conclude that there is any correlation between.
-   With this information, we may conclude that countries in Africa with lower and higher development tend to be equally far away from being fully green, and targetting either group specifically for increasing their renewable capabilities is not useful.

## Question #3: \[Hyman - Private Sector Participation and Renewable Energy Adoption\]

Is private sector participation in electricity generation capacity associated with renewable energy adoption in African countries?

-   Electricity is essential, but countries organize their power sectors differently.
-   Some rely more on private sector participation, while others rely more on public control.
-   This makes it important to examine whether private sector involvement is related to renewable energy adoption.
-   This connects directly to our theme of renewables, since the structure of the electricity sector may influence how easily countries transition to cleaner energy.
-   Understanding this relationship can help inform future energy and climate policy.

## Q3: Variables & Approach

**Variables used:**

-   **Private Sector Power Generation:**: the amount of installed electricity generation capacity that is commercially operated by private companies.
-   **Renewables Share of Generation Mix (%):** the share of total electricity generation that comes from renewable sources. 
-   **Installed Generation Capacity:** the total maximum generating capacity of a country’s power system.

**New variables/groups defined:**

-   **Private Generation Share**: the proportion of a country’s total installed electricity generation capacity that is operated by the private sector. It is calculated by dividing commercially operating private-sector generation capacity by total installed generation capacity, then multiplying by 100.

## Q3: Visualization

```{r q3-plot}
CATF_clean <- read_csv("CATF_data.csv") |>
  filter(
    !is.na(country),
    !is.na(private_sector_power_gen_commercial),
    !is.na(installed_gen_capacity),
    !is.na(renewables_share_gen_mix_percent),
    installed_gen_capacity > 0,
    renewables_share_gen_mix_percent >= 0,
    renewables_share_gen_mix_percent <= 100
  ) |>
  mutate(
    private_generation_share = (private_sector_power_gen_commercial / installed_gen_capacity) * 100
  ) |>
  filter(
    private_generation_share >= 0,
    private_generation_share <= 100
  ) |>
  group_by(country) |>
  summarise(
    private_generation_share = median(private_generation_share),
    renewables_share_gen_mix_percent = median(renewables_share_gen_mix_percent)
  ) |>
  mutate(
    private_group = if_else(
      private_generation_share < 50,
      "Public-oriented",
      "Private-oriented"
    )
  )

ggplot(
  CATF_clean,
  aes(x = private_generation_share, y = renewables_share_gen_mix_percent)
) +
  geom_point(size = 3, color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.9) +
  labs(
    title = "Renewable Energy Share vs Private Sector Participation",
    x = "Private share of installed generation capacity (%)",
    y = "Renewables share of generation mix (%)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

## Q3: Summary Table

```{r q3-table}
q3_summary <- CATF_clean |>
  group_by(private_group) |>
  summarise(
    Countries = n(),
    `Mean renewable share (%)` = round(mean(renewables_share_gen_mix_percent), 1),
    `Median renewable share (%)` = round(median(renewables_share_gen_mix_percent), 1)
  )

knitr::kable(
  q3_summary,
  caption = "Renewable energy share by electricity system orientation"
)
```

The table compares public-oriented and private-oriented electricity systems by showing the number of countries in each group, along with the mean and median renewable share. Public-oriented countries have less than 50% private generation capacity, while private-oriented countries have 50% or more.

## Insights for Q3

-   The scatterplot suggests a weak negative relationship between private sector participation and renewable energy share.
-   Public-oriented countries appear to have higher average and median renewable shares than private-oriented countries.
-   However, there is still considerable variation across countries, indicating that other factors likely also influence renewable energy adoption.
-   The group comparison should be interpreted cautiously because the private-oriented group is much smaller than the public-oriented group.
-   Overall, the analysis suggests a weak association rather than a strong conclusion, and it should not be interpreted as evidence of causation.

## Question #4 (Sahand Samadirad): Revenue Collection and Renewable Capacity

**Do countries with higher electricity revenue collection rates tend to have more installed renewable energy capacity?**

- Revenue collection reflects the **financial health of utilities**.
- Stronger finances can support **renewable energy investment**.
- Weak collection may slow infrastructure expansion.

- This matters to **CATF** because utility finances may affect energy transition progress.
- It connects to our theme by testing whether **financial performance** is associated with renewable energy development.

**Hypothesis:**  
Countries with higher electricity revenue collection rates will tend to have **greater installed renewable energy capacity**.

## Q4: Variables & Approach

**Variables used:**

- **Electricity revenue collection rate (%)**  
  Measures how much billed electricity revenue utilities successfully collect.

- **Installed renewable energy capacity (MW)**  
  Measures the scale of solar and wind development in each country.

**Grouping used:**

Countries were grouped by revenue collection performance:

- **Low:** below 70%
- **Medium:** 70–90%
- **High:** above 90%

This allows a clearer comparison of renewable capacity across different levels of financial performance.

## Q4: Visualization

```{r q4-plot, fig.width=7.2, fig.height=4.2, out.width="78%"}
CATF_data |>
  group_by(country) |>
  summarise(
    collection_rate_percent = median(collection_rate_percent, na.rm = TRUE),
    total_installed_variable_renewables_capacity = median(total_installed_variable_renewables_capacity, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(
    !is.na(collection_rate_percent),
    !is.na(total_installed_variable_renewables_capacity)
  ) |>
  ggplot(aes(x = collection_rate_percent, y = total_installed_variable_renewables_capacity)) +
  geom_point(color = "darkgreen", size = 2.6) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  scale_y_log10() +
  labs(
    x = "Revenue Collection Rate (%)",
    y = "Renewable Capacity (MW)",
    title = "Revenue Collection and Renewable Capacity Across Countries"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16),
    axis.title = element_text(size = 13),
    axis.text = element_text(size = 11),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey80")
  )
```
\vfill
{\footnotesize The y-axis uses a logarithmic scale to make cross-country differences easier to see.}

## Q4: Summary Table

```{r q4-table, out.width="75%"}
CATF_data |>
  group_by(country) |>
  summarise(
    collection_rate_percent = median(collection_rate_percent, na.rm = TRUE),
    total_installed_variable_renewables_capacity = median(total_installed_variable_renewables_capacity, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(
    !is.na(collection_rate_percent),
    !is.na(total_installed_variable_renewables_capacity)
  ) |>
  mutate(
    collection_group = case_when(
      collection_rate_percent < 70 ~ "Low (<70%)",
      collection_rate_percent < 90 ~ "Medium (70–90%)",
      TRUE ~ "High (>90%)"
    )
  ) |>
  mutate(
    collection_group = factor(
      collection_group,
      levels = c("Low (<70%)", "Medium (70–90%)", "High (>90%)")
    )
  ) |>
  group_by(collection_group) |>
  summarise(
    Countries = n(),
    `Avg Rate (%)` = round(mean(collection_rate_percent, na.rm = TRUE), 1),
    `SD Rate` = round(sd(collection_rate_percent, na.rm = TRUE), 1),
    `Avg Capacity (MW)` = round(mean(total_installed_variable_renewables_capacity, na.rm = TRUE), 1),
    `SD Capacity` = round(sd(total_installed_variable_renewables_capacity, na.rm = TRUE), 1),
    .groups = "drop"
  ) |>
  kable(caption = "Renewable Capacity by Collection Rate Category")
```

- High-collection countries have the highest average renewable capacity.
- However, variation is large within groups, especially in the high category.

## Insights for Q4

- There is a **positive relationship** between revenue collection and renewable capacity: countries in the high collection group (~970 MW) have much higher average capacity than those in medium (~250 MW) and low (~106 MW) groups.
- However, there is **substantial variation within groups**, especially among high-collection countries, indicating that countries with similar collection rates can still have very different renewable capacities.
- The regression line has a **slight positive slope**, but the wide variability of data points indicates that collection rate is only able to explain a limited portion of the variation.
- Overall, while stronger revenue collection is linked to a higher renewable capacity, it cannot serve as a strong predictor by itself; other factors tend to have more influence.

## Overall Limitations

Data Limitations:

-   **Data Collection:** Because the data set combines information from various global databases and public records, data for different countries often come from different reporting years or rely on slightly different definitions and metrics. These differences could introduce inaccuracy during data analysis.

-   **Unmeasured Factors:** Our analysis focused on relationships between two variables at a time, meaning we were unable to control for external factors. Hidden third variables such as foreign aid investments, geographic solar/wind potential and political stability could be responsible for the outcomes of both variables.

-   **Missing Data:** Several countries had incomplete reports for variables (such as SAIFI or private sector generation capacity). Omitting these missing values reduces the overall sample size, meaning our findings may be less representative of the African continent.

## Overall Conclusion

- Across all research questions, the results show that renewable energy development in African electricity systems is influenced by **multiple interacting factors**, and not just a single one.

- Financial performance (revenue collection), development level, and private sector participation show **some association with renewable outcomes**, but individually, they do not strongly explain differences across countries.

- In particular, while higher revenue collection is associated with higher renewable capacity, there is only a **weak and highly variable** relationship, indicating that other factors such as policy, infrastructure, and country size could play bigger roles.

- Overall, this supports our theme that renewable energy development is shaped by a **combination of financial, institutional, and structural conditions**, rather than any single metric.

- A key limitation is that the dataset contains **missing and inconsistent values**, and  hence, the analysis utilises aggregated country-level data, which may hide within-country variation and introduce selection bias.

- For CATF, this suggests that improving renewable energy outcomes will likely require **coordination across different fields**, rather than focusing on a single performance indicator.