---
title: "Renewable Energy Development and Electricity System Performance in Africa"
subtitle: "An Exploratory Analysis Using the CATF Electricity Sector Dataset"
author: "Caellum Yip Hoi Lee | Paanini Kartik | Hyman Al-Qaissi | Sahand Samadirad | \n Group C"
date: today
format:
  beamer:
    theme: Madrid
    colortheme: default
    fonttheme: serif
    aspectratio: 169
    slide-level: 2
    toc: false
    incremental: false
    pdf-engine: pdflatex
    keep-tex: false
editor: visual
---

```{r setup, include=FALSE}
library(gridExtra)
library(tidyverse)
library(knitr)
library(cowplot)
library(broom)
knitr::opts_chunk$set(
  echo = FALSE,
  include = TRUE,
  warning = FALSE,
  message = FALSE,
  results = "markup",
  fig.width = 6,
  fig.height = 3.5,
  fig.align = "center",
  out.width = "80%"
)
```

```{r load-data, echo=FALSE}
### Read in data
CATF_data <- read_csv("CATF_data.csv")
```

Introduction

Overall Theme: Patterns and drivers of renewable energy development across African electricity systems

Research Goal: Investigate how renewable energy usage is different among African electricity systems and explore factors that may be associated with differences in renewable energy capacity between countries.

Target Population: The study focuses on African countries included in the CATF electricity sector dataset.

Audience: The primary audience is Clean Air Task Force (CATF) staff, who are interested in understanding patterns in electricity sector performance across African countries to help inform the development of the Electricity Sector Performance Scorecard (CATF-ESPS).

Data Summary

The dataset contains country-level indicators describing electricity system performance across African countries. These indicators capture areas such as electricity access, reliability, infrastructure development, financial performance, and renewable energy capacity.

Data Wrangling:

\- To facilitate comparison across countries, the variable `collection_rate_percent` was grouped into categories representing different levels of electricity revenue collection performance. A new variable called `collection_group` was created in the `CATF_data` dataset. The groups are defined as Low (\<70%), Medium (70–90%), and High (\>90%). This transformation allows us to compare `total_installed_variable_renewables_capacity` across countries with different levels of revenue collection performance.

## Question #1: \[Caellum Yip Hoi - Lee - Renewable Presence and Power Reliability\]

Do countries with a higher share of renewable energy in the generation mix tend to experience more frequent power interruptions?

* **Relevance:** As African nations develop their grids, transitioning to renewable energy is a priority. Understanding how the shift to renewables could impact energy reliability may be helpful in leading responsible development. 
* **Connection to Theme:** This investigates a potential trade-off between two core areas of the CATF-ESPS: renewable energy development and power grid stability.

## Q1: Variables & Approach

**Variables used:**

-   **renewables_share_gen_mix_percent:** The percentage of electricity generation coming from renewable energy sources. This variable captures the amount of renewable energy being employed in the country's electricity system.

-   **SAIFI:** System Average Interruption Frequency Index, representing Voltage Quality / Frequency It is the total number of reported interruptions divided by total customers. This variable serves as a primary measure of power reliability.

**New variables/groups defined:**

To intuitively compare reliability, I can group countries based on their renewable generation share:

-   renewable_group: "Low" when renewables_share_gen_mix_percent is below 20%, "Medium" when it is between 20-50%, and "High" when above 50%

## Q1: Visualization

```{r q1-plot}
CATF_data |>
  # An extreme outlier of 65,000 % exists in the data, filtered this out because it makes no sense and is likely a typo. 
  filter(renewables_share_gen_mix_percent <= 100) |> 
  ggplot(aes(x = renewables_share_gen_mix_percent, y = SAIFI)) +
  geom_point(color = "steelblue", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    x = "Renewables Share in Generation Mix (%)",
    y = "Frequency of Interruptions (SAIFI)",
    title = "Renewable Energy Share vs Power Reliability"
  )
```

## Q1: Summary Table

```{r q1-table}
CATF_data |>
  # An extreme outlier of 65,000 % exists in the data, filtered this out because it makes no sense and is likely a typo. 
  filter(renewables_share_gen_mix_percent <= 100) |>
  mutate(
    renewable_group = case_when(
      renewables_share_gen_mix_percent < 20 ~ "Low (<20%)",
      renewables_share_gen_mix_percent <= 50 & renewables_share_gen_mix_percent >= 20 ~ "Medium (20–50%)",
      renewables_share_gen_mix_percent > 50  ~ "High (>50%)"
    )
  ) |>
  
  mutate(renewable_group = factor(renewable_group, levels = c("Low (<20%)", "Medium (20–50%)", "High (>50%)"))) |>
  group_by(renewable_group) |>
  summarise(
    Countries = n_distinct(country),
    Observations = n(),
    `Avg Renewable Share (%)` = round(mean(renewables_share_gen_mix_percent, na.rm = TRUE), 1),
    `Avg SAIFI (Interruptions)` = round(mean(SAIFI, na.rm = TRUE), 1)
  ) |>
  kable(caption = "Average System Interruptions (SAIFI) by Renewable Energy Share")
```

## Q1: Insights & Takeaways

-   **No clear linear relationship:** The scatter plot has a nearly flat trend line, suggesting that a higher share of renewables in the generation mix is not correlated with a higher frequency of power interruptions.
-   **Reliability driven by other factors:** Countries with high interruption frequencies had widely varying levels of renewable energy in their generation mix. This suggests that issues with reliability and the subsequent SAIFI score likely has more to do with other factors, i.e infrastructure.
-   **Renewables not sacrificing stability:** I see this evidence as positive, since it suggests that African electricity systems can have higher percentages of renewable energy without directly inducing a lower electric grid reliability.

## Question #2: \[Paanini Kartik - Development and Renewable Energy Usage\]

**Do more developed countries tend to generate a higher or lower proportion of renewable energy?**

-  From one perspective, we may expect to see developed countries have a **higher proportion** of their energy generation capabilities come from renewable energy, given there is more capital available to invest and higher levels of education could mean more skilled workers capable to building renewable energy sources.

-  From another perspective, we may expect to see developed countries have a **lower share in renewables**, as their higher development may correlate with the choice to use cheaper sources of energy such as oil and coal.

-  Understanding how development correlates with renewables is crucial to understanding electricty systems in Africa and how renewable usage is distributed in the continent.

## Q2: Variables & Approach

**Variables from data:**

-   **hdi** The Human Development Index of a given country. This will be used as a metric for how developed the country is, where HDI measures factors such as life expectancy, education, and GNI to assess how developed a country is.
-   **renewables_share_gen_mix_percent** What % of the country's generation comes from renewables. We use this instead of total_installed_renewables_capacity as that may trivially become higher with more development.

**Created variables:**

We will create a new variable by grouping countries based on their HDI using the standard detailed by the UNDP, which defines: "Low HDI" as less than 0.5, "Medium HDI" as between 0.5 and 0.699, "High HDI" as between 0.7 and 0.799, and "Very High HDI" as 0.8 and above.

## Q2: Visualization

```{r q2-plot}
CATF_data |>
  filter(!is.na(hdi), !is.na(renewables_share_gen_mix_percent)) |>
  filter(renewables_share_gen_mix_percent <= 100 &
         renewables_share_gen_mix_percent >= 0) |>
  filter(hdi >= 0 & hdi <= 1) |> # remove invalid values
  ggplot(aes(x = hdi, y = renewables_share_gen_mix_percent)) +
  geom_point(color = "darkred", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    x = "Human Development Index (HDI)",
    y = "Renewables Share of Generation Mix (%)",
    title = "Development Level vs Renewable Energy Usage"
  )
```

## Q2: Summary Table

```{r q2-table}
CATF_data |>
  filter(!is.na(hdi), !is.na(renewables_share_gen_mix_percent)) |>
  filter(renewables_share_gen_mix_percent <= 100 &
         renewables_share_gen_mix_percent >= 0) |>
  filter(hdi >= 0 & hdi <= 1) |> # remove invalid values
  mutate(
    hdi_grouping = case_when(
      hdi < 0.5 ~ "Low (<0.5)",
      hdi >= 0.5 & hdi < 0.7 ~ "Med. (0.5–0.7)",
      hdi >= 0.7 & hdi < 0.8 ~ "High (0.7–0.8)",
      hdi >= 0.8 ~ "Very High (>=0.8)"
    )
  ) |>
  mutate(
    hdi_grouping = factor(
      hdi_grouping,
      levels = c("Low (<0.5)", "Med. (0.5–0.7)", "High (0.7–0.8)", "Very High (>=0.8)")
    )
  ) |>
  group_by(hdi_grouping) |>
  summarise(
  Countries = n(),
    `Avg HDI` = round(mean(hdi, na.rm = TRUE), 2),
    `HDI SD` = round(sd(hdi, na.rm = TRUE), 2),
    `Avg Renewable Share (%)` = round(mean(renewables_share_gen_mix_percent, na.rm = TRUE), 1),
    `Renewable Share SD` = round(sd(renewables_share_gen_mix_percent, na.rm = TRUE), 1)
) |> rename(`HDI Group` = hdi_grouping) |>
     kable(caption = "Renewable Energy Share by HDI Development Group")
```

## Q2: Linear Model Output

```{r q2-lm}
lm_q2 <- CATF_data |>
  filter(!is.na(hdi), !is.na(renewables_share_gen_mix_percent)) |>
  filter(renewables_share_gen_mix_percent <= 100 &
         renewables_share_gen_mix_percent >= 0) |>
  filter(hdi >= 0 & hdi <= 1) |>
  lm(renewables_share_gen_mix_percent ~ hdi, data = _) |>
  tidy()

tibble(
  Term = c("Intercept", "Slope (HDI)"),
  Estimate = c(
    lm_q2 |> filter(term == "(Intercept)") |> select(estimate) |> pull(),
    lm_q2 |> filter(term == "hdi") |> select(estimate) |> pull()
  ),
  `p-Value` = c(
    lm_q2 |> filter(term == "(Intercept)") |> select(p.value) |> pull(),
    lm_q2 |> filter(term == "hdi") |> select(p.value) |> pull()
  )
) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable(caption = "Linear Model: Renewables Share ~ HDI", align = "c")
```

## Q2: Insights & Takeaways

-   Visually speaking, there appears to be a **slight negative correlation** between the two variables, given the line of best fit having a slightly negative slope.
-   However, there appears to be a large spread in the data, with data points not clustering around said line. The spread is more pronounced for countries with lower HDI scores, with their standard deviations being around 30, while countries with higher HDI scores have standard deviations at around 20 or less.
-   Despite this, the **p-value of 0.001 and 0.000 for the slope and intercept respectively** are well below the common alpha threshold of 0.05, which suggests that the correlation is definite and **we have strong evidence to reject the null hypothesis**. Thus, we can conclude there is a negative correlation between how developed a country is and what % of its energy comes from renewables.
-   With this information, it would be beneficial to understand why this negative correlation exists so as to eliminate the gap and encourage more renewable usage in the more developed countries in Africa.

## Question #3: \[Hyman - Private Sector Participation and Renewable Energy Adoption\]

Is private sector participation in power generation associated with renewable energy adoption in African countries?

## Q3: Variables & Approach

**Variables used:**

-   private_sector_power_gen_commercial
-   renewables_share_gen_mix_percent
-   installed_gen_capacity

**New variables/groups defined:**

-   private_generation_share \<- private_sector_power_gen_commercial / installed_gen_capacity

## Q3: Visualization

```{r q3-plot}
CATF_clean <- read_csv("CATF_data.csv")|>
  filter(
    !is.na(private_sector_power_gen_commercial),
    !is.na(installed_gen_capacity),
    !is.na(renewables_share_gen_mix_percent),
    installed_gen_capacity > 0
  ) |>
  mutate(
    private_generation_share = private_sector_power_gen_commercial / installed_gen_capacity
  )

q3_plot <- ggplot(CATF_clean, aes(x = private_generation_share, y = renewables_share_gen_mix_percent)) +
  geom_point() +
  labs(
    title = "Renewable Share vs Private Generation Share",
    x = "Private generation share",
    y = "Renewables share of generation mix (%)"
  )

q3_plot
```

## Q3: Summary Table

```{r q3-table}
q3_summary <- CATF_clean |>
  mutate(
    private_group = if_else(private_generation_share < 0.5, "Lower", "Higher")
  ) |>
  group_by(private_group) |>
  summarise(
    n = n(),
    mean_renewable = mean(renewables_share_gen_mix_percent),
    median_renewable = median(renewables_share_gen_mix_percent)
  )

q3_summary
```

## Insights for Q3

-   There is no clear and strong relationship between private sector participation in power generation and renewable usage.

-   Countires with lower private participation have a variety of renewable outcomes, whilst countries with higher private participation did not have higher renewable outcomes.

-   The median across both higher and lower groups was the same, further suggesting that renewable usage may not be explained solely by the private sectors role in power generation.

## Question #4: [Sahand Samadirad - Revenue Collection and Renewable Capacity]

**Do countries with higher electricity revenue collection rates tend to have more installed renewable energy capacity?**

- Electricity revenue collection reflects the **financial health of utilities** and their ability to fund infrastructure.
- Renewable energy projects require **high upfront investment**, making financing a key constraint.
- Weak collection rates may limit investment, leading to **slower renewable energy development**.

- This directly connects to our theme: identifying **drivers of renewable energy development in African electricity systems**.
- Specifically, it tests whether **financial performance (collection rate)** influences renewable capacity expansion.

**Hypothesis:**  
Countries with higher electricity revenue collection rates will tend to have **greater installed renewable energy capacity** due to stronger financial capacity for investment.

## Q4: Variables & Approach

**Variables used:**

-   **collection_rate_percent**\
    Percentage of billed electricity revenue that utilities successfully collect.\
    This variable reflects the financial efficiency and operational performance of electricity utilities.

-   **total_installed_variable_renewables_capacity**\
    Total installed capacity of variable renewable energy sources such as wind and solar (MW).\
    This variable represents the level of renewable energy development within a country's electricity system.

**New variables/groups defined:**

To make comparisons clearer, countries are grouped based on electricity revenue collection performance:

-   Low collection rate: below 70%\
-   Medium collection rate: 70–90%\
-   High collection rate: above 90%

These categories allow us to compare renewable capacity across different levels of financial performance.

## Q4: Visualization

```{r q4-plot}
CATF_data |>
  ggplot(aes(x = collection_rate_percent, y = total_installed_variable_renewables_capacity)) +
  geom_point(color = "darkgreen", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  scale_y_log10() +
  labs(
    x = "Electricity Revenue Collection Rate (%)",
    y = "Installed Renewable Energy Capacity (MW)",
    title = "Collection Rate vs Renewable Energy Capacity"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),   # removes dense log gridlines
    panel.grid.major = element_line(color = "grey80")  # keeps clean standard grid
  )
```

The y-axis is shown on a logarithmic scale to account for large differences in renewable capacity across countries.

## Q4: Summary Table

```{r q4-table}
CATF_data |>
  filter(
    !is.na(collection_rate_percent),
    !is.na(total_installed_variable_renewables_capacity)
  ) |>
  mutate(
    collection_group = case_when(
      collection_rate_percent < 70 ~ "Low (<70%)",
      collection_rate_percent < 90 ~ "Medium (70–90%)",
      TRUE ~ "High (>90%)"
    )
  ) |>
  mutate(
    collection_group = factor(
      collection_group,
      levels = c("Low (<70%)", "Medium (70–90%)", "High (>90%)")
    )
  ) |>
  group_by(collection_group) |>
  summarise(
    Countries = n(),
    `Avg Collection Rate (%)` = round(mean(collection_rate_percent, na.rm = TRUE), 1),
    `SD Collection Rate (%)` = round(sd(collection_rate_percent, na.rm = TRUE), 1),
    `Avg Renewable Capacity (MW)` = round(mean(total_installed_variable_renewables_capacity, na.rm = TRUE), 1),
    `SD Renewable Capacity (MW)` = round(sd(total_installed_variable_renewables_capacity, na.rm = TRUE), 1)
  ) |>
  kable(caption = "Renewable Capacity by Collection Rate Category")
```

## Insights for Q4

- The scatterplot and regression line suggest a **weak positive relationship** between electricity revenue collection rate and installed renewable energy capacity.
- Countries with higher collection rates tend to have **higher average renewable capacity**, but there is **substantial variation within each group**, particularly in the high-collection category.
- The fitted linear regression line has a **slightly positive slope**, but the wide spread of data points indicates **low explanatory power**, meaning the model does not explain much of the variation in renewable capacity.
- Some collection rates exceed 100%, which may occur due to the recovery of past unpaid bills (arrears) or differences in reporting methods across countries. These values increase variability, especially in the high-collection group, and can influence the averages and spread of the data.
- This analysis uses only observations with complete data for both variables, ensuring a **consistent and comparable sample** when examining the relationship. While this improves the validity of the comparison, it may introduce **selection bias**, as countries with missing renewable capacity data are excluded and may differ systematically from those included.
- Overall, this suggests that while financial performance may play a role, **collection rate alone is not a strong predictor** of renewable energy development, and other factors such as country size, economic conditions, and policy likely have a greater impact.
- In summary, there is only weak evidence that higher collection rates are associated with greater renewable energy capacity, and the relationship is not strong enough to suggest that collection rate is a key driver on its own.

## Overall Limitations

Data Limitations:

-   **Data Collection:** The CATF data was collected by student internet searches. This introduces lots of room for inaccuracies, since data points may come from different years or use different definitions.

-   **Inaccuracies in data:** Outliers such as the 65 000% renewables_share_gen_mix indicate that the data set contains some inaccurate data that will impact the results of the research.

-   **Missing Data:** Many African countries may lack the necessary infrastructure to report all the variables in the dataset. This may skew the results of research; it may cause the less wealthy countries to be less represented in the dataset.

## Future Directions

-   **Standardized Data Collection:** To reduce inaccuracies introduced by varying methods of data collection, the CATF dataset could be collected only from standardized global databases, i.e World Bank. This would minimize inconsistent value but may lead to more values being dropped.
-   **Data Cleansing:** To improve the removal of extreme outliers (like the 65,000% renewable share), strict checks could be implemented during data entry to automatically remove physically impossible values before analysis. For example, a check could be implemented to ensure that the values inputted in the renewables_share_gen_mix are never greater than 100.
-   **Inclusive Data Collection Methods:** Instead of collecting data directly from the government records, globally available proxy data could be used to ensure that every country is included. While it is unfeasible to manually collect the data for each country, the CATF could use stand-in values as estimates. For example, the amount of imported solar panels could be used as a rough proxy for solar capacity.
